---
title: "Assessment"
author: "Naveed"
date: "2024-11-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Research Question

How does the prescription of weight loss drugs (specifically GLP-1 antagonists) vary across socioeconomically diverse regions in Scotland in 2023?


## The Libraries

This section contains the code for loading in all the required libraries. 

```{r}
library(tidyverse)
library(here)
library(janitor)
library(gt)
library(sf)
library(ggspatial)
```

## The Data

This section contains the code for loading in all the required datasets. 

```{r pressure, echo=FALSE}
files <- list.files(here("data", "year_data"), pattern = "csv", full.names = TRUE)

year_data <- files %>%
  map_dfr(~read_csv(., col_types = cols(
    DMDCode = col_character()
  ))) %>%
  clean_names()

hb_data <- read_csv("https://www.opendata.nhs.scot/dataset/9f942fdb-e59e-44f5-b534-d6e17229cc7b/resource/652ff726-e676-4a20-abda-435b98dd7bdc/download/hb14_hb19.csv") %>% 
  clean_names()

special_hb_data <- read_csv("https://www.opendata.nhs.scot/dataset/65402d20-f0f1-4cee-a4f9-a960ca560444/resource/0450a5a2-f600-4569-a9ae-5d6317141899/download/special-health-boards_19022021.csv") %>% 
  clean_names()

deprivation_data <- read_csv("https://www.opendata.nhs.scot/dataset/78d41fa9-1a62-4f7b-9edb-3e8522a93378/resource/acade396-8430-4b34-895a-b3e757fa346e/download/simd2020v2_22062020.csv") %>% 
  select(HB, SIMD2020V2CountryDecile) %>% 
  clean_names()

population_data <- read_csv(here( "data", "UV103_age_health_board_census.csv"), skip = 10) %>% 
  rename(Spare = "...6",
         hb_name = "Health Board Area 2019",
         hb_population = Count) %>% 
  filter(Age == "All people" & Sex == "All people") %>% 
  select(hb_name, hb_population) %>% 
  mutate(hb_name = paste("NHS", hb_name)) %>% 
    clean_names()
  
  NHS_healthboards <- st_read(here("data", "NHS_healthboards_2019.shp"))
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

This section of code joins the various datasets of interest to create the joined data.
It then filters based on the relevant requirements.  
```{r}
weightdrug_data <- right_join(hb_data, year_data, by = join_by(hb == hbt)) %>% 
  filter(str_detect(bnf_item_description, "GLUTIDE|OZEMPIC|SAXENDA|TRULICITY")) %>% 
  group_by(hb_name, hb, bnf_item_description) %>% 
  summarise(paid_quantity = sum(paid_quantity))

joined_data <- left_join(weightdrug_data, population_data) %>%
  left_join(., NHS_healthboards, by = join_by(hb == HBCode))

joined_data <- left_join(joined_data, deprivation_data, by = join_by(hb == hb))
```

This section of code generates a graph.

First graph shows the comparison between SIMD and prescriptions of weigh loss drugs.

The second and third graph show the distribution of these prescriptions around Scotland.

I would like to explore the relationship between BMI and deprivaton and how that affects prescription of weight loss drugs.


```{r}
joined_data %>% 
ggplot(aes(x = paid_quantity, y = simd2020v2country_decile)) +
  geom_smooth() +
  theme(axis.text.x = element_text( 
    angle = 90,
    size=7, 
    face=3)) +
  labs(title = "Prescriptions vs SIMD", x = "Prescriptions", y = "SIMD")

weightdrug_data %>% 
  ggplot(aes(y = paid_quantity, x = reorder(hb_name, paid_quantity))) +
  geom_col() +
  coord_flip()+
  labs(title = "No. of weight loss drug prescriptions per health board", x = "Health Board Name", y = "Weight Loss Drugs Prescriptions per person")

joined_data %>% 
ggplot(aes(geometry = geometry, fill = paid_quantity)) +
  geom_sf() +
  scale_fill_distiller(palette = "GnBu") +
  labs(title = "Number of Weight Loss Drug Prescriptions", subtitle = "In Scotland", fill = "Total No. of Prescriptions") +  
  annotation_scale(location = "tl") +
  annotation_north_arrow(
    location = "tl",    
    pad_y = unit(0.5, "in"),    
    style = north_arrow_nautical(
      fill = c("blue", "white"),      
      line_col = "grey20"
    ) 
  )
```

